# Simulations

```{r}
source("helper.R")
```

In order to exemplify the utility of this package, a handful of hypothetical
data will be presented, together with a proposal of an appropriate likelihood
function that is expected to segment the data in an expected manner.

## Correlated random variables

Let $C_1$, $C_2$, ..., $C_n$ a sequence of $n$ vectors of data points, each with $m$
observations represented by $C_k = \{c_{k1},\ c_{k2},\ ...,\ c_{km}\}$, with
$1 \le k \le n$. Assume that each $C_k$ is a function of an unknown set of independent
random variables $X$, such that $C_k = f_k(X)$. When used together with a multivariate
likelihood [@park2017fundamentals], for which an implementation is provided in
the `segmentr` package as the `multivariate()` function, it's possible to identify
the set of $N_b$ change points
$b_1, ..., b_{N_b}$, with $b_0 = 1$ and $b_{N_b + 1} = n + 1$, such that all $f_k$, for
$b_i \le k \le b_{i+1}-1$ is a function of the same set of independent random variables $X$,
or each $1 \le i \le N_b$.

Let $X_1, ..., X_6$ be independent random variables and $C_1, ..., C_{15}$ be
defined as in \@ref(eq:examplecolumns) . It's possible to see the first segment $C_1, ..., C_5$
depend on $X_1, X_2$, the second segment $C_6, ..., C_{10}$ depend on $X_3, X_4$ and the third
and last segment $C_{11}, ..., C_{15}$ depend on $X_4, X_5$. Therefore, the change points would
be $b_1 = 6$ and $b_2 = 11$.

\begin{equation}
\begin{aligned}
C_1 & = X_1 \\
C_2 & = X_1 - X_2 \\
C_3 & = X_2 \\
C_4 & = X_1 + X_2 \\
C_5 & = X_1 \\
C_6 & = X_3 \\
C_7 & = X_3 - X_4 \\
C_8 & = X_4 \\
C_9 & = X_3 + X_4 \\
C_{10} & = X_3 \\
C_{11} & = X_5 \\
C_{12} & = X_5 - X_6 \\
C_{13} & = X_6 \\
C_{14} & = X_5 + X_6 \\
C_{15} & = X_5
\end{aligned}
(\#eq:examplecolumns)
\end{equation}

To illustrate, consider the following code, which defines $X_1, ..., X_6$ and $C_1, ..., C_{15}$ as specified in \@ref(eq:examplecolumns), and also consider the matrix $D$ as the column binding of the vectors $C_1, ..., C_{15}$.

(ref:sample-columns-caption) Sample values of the correlated random variables defined in \@ref(eq:examplecolumns).

```{r sample-columns}
n <- 100
X1 <- sample(1:2, n, replace = TRUE)
X2 <- sample(1:2, n, replace = TRUE)
X3 <- sample(1:2, n, replace = TRUE)
X4 <- sample(1:2, n, replace = TRUE)
X5 <- sample(1:2, n, replace = TRUE)
X6 <- sample(1:2, n, replace = TRUE)

C1 <- X1
C2 <- X1 - X2
C3 <- X2
C4 <- X1 + X2
C5 <- X1
C6 <- X3
C7 <- X3 - X4
C8 <- X4
C9 <- X3 + X4
C10 <- X3
C11 <- X5
C12 <- X5 - X6
C13 <- X6
C14 <- X5 + X6
C15 <- X5
D_example <- cbind(C1, C2, C3, C4, C5, C6, C7, C8,
           C9, C10, C11, C12, C13, C14, C15)
head(D_example) %>%
  kable(caption="Sample values of the correlated random
        variables defined in (ref:eq-examplecolumns).")
```

It's possible to identify the segments in the matrix $D$ with a penalized version of the
`multivariate` function included in the package, as shown in Table \@ref(tab:results-multivariate-penalized)


(ref:eq-examplecolumns) \@ref(eq:examplecolumns)

```{r results-multivariate-penalized}
penalized_multivariate = function(X) multivariate(X) - 2 ^ ncol(X)

results_multivariate_penalized <- segment(
  D_example,
  likelihood = penalized_multivariate,
  algorithm = "exact"
)

print_results_table(
  results_multivariate_penalized,
  caption="Segment tables as defined in
  (ref:eq-examplecolumns)."
)
```

Notice it's important to use a penalized likelihood version of the `multivariate` function because
the original function favors bigger segments. The answer provided by the `segment` package
in that scenario would actually be all the columns in the $D$ matrix, as shown in Table \@ref(tab:results-multivariate-non-penalized).

```{r results-multivariate-non-penalized}
segment(D_example, likelihood = multivariate, algorithm = "exact") %>%
  print_results_table(
    caption="Results of segmentation usign a non-penalized
      multivariate likelihood")
```

## Segments with similar averages

[@homozygosity] describes a process on how to find windows of
contiguous homozygosity, i.e. segments in the genetic data in which the alleles
are of the same type. This is of interest for a researching investigating
diseases. Considering this problems scenario, Segmentr can be used to segment
random variables $C_1, ..., C_n$ that represent genetic data, encoded as zero
for homozygosity and one for heterozygosity, i.e. zero when the alleles are of
the same type and one when the alleles are different.

So, in order to use Segmentr to solve this problem, it's necessary
to find a likelihood function that favors segments the homogeneity
of a given segment. That problem is approached by proposing a
"mean" likelihood function, i.e. a function that maximizes the
likelihood of segments whose elements approximate the segment
average. One such function is defined in
\@ref(eq:mean-likelihood). Notice, however, the likelihood
function proposed favors a single column segments, as it's logically
the entity whose elements most approximate the average of the
entire window. In order to counter this undesirable behavior, we
penalized the function by subtracting from it a single constant,
as described in \@ref(eq:penalizedmeanlikelihood). The constant
factor has the effect of adding up when too many segments are
considered in the estimation process, making it so wider segments
end being picked up in the estimation process.

In order to observe how the proposed function behaves, consider the simple
example defined in \@ref(eq:geneticexample), in which $X_i$ for $i \in \{1,
\dots, 20\}$ represent each a column indexed by $i$ of a data set $D$, and
$\text{Bern(p)}$ representing the Bernoulli distribution with probability $p$.
When segmenting the data set defined in \@ref(eq:geneticexample) with the
likelihood function defined in \@ref(eq:penalizedmeanlikelihood), we obtain the
results displayed in Table \@ref(tab:results-penalized-mean).

\begin{equation}
L_{\mu}(X)=-\sum_i(x_i-E[X])^2 \;\bigg|\; x_i \in X
(\#eq:mean-likelihood)
\end{equation}

\begin{equation}
P_{L_{\mu}}(X)=L_{\mu}(X)-1
(\#eq:penalizedmeanlikelihood)
\end{equation}

\begin{equation}
\begin{aligned}
X_{1} & \sim \text{Bern}(0.9) \\
X_{2} & \sim \text{Bern}(0.9) \\
X_{3} & \sim \text{Bern}(0.9) \\
X_{4} & \sim \text{Bern}(0.9) \\
X_{5} & \sim \text{Bern}(0.9) \\
X_{6} & \sim \text{Bern}(0.1) \\
X_{7} & \sim \text{Bern}(0.1) \\
X_{8} & \sim \text{Bern}(0.1) \\
X_{9} & \sim \text{Bern}(0.1) \\
X_{10} & \sim \text{Bern}(0.1) \\
X_{11} & \sim \text{Bern}(0.1) \\
X_{12} & \sim \text{Bern}(0.1) \\
X_{13} & \sim \text{Bern}(0.1) \\
X_{14} & \sim \text{Bern}(0.1) \\
X_{15} & \sim \text{Bern}(0.1) \\
X_{16} & \sim \text{Bern}(0.9) \\
X_{17} & \sim \text{Bern}(0.9) \\
X_{18} & \sim \text{Bern}(0.9) \\
X_{19} & \sim \text{Bern}(0.9) \\
X_{20} & \sim \text{Bern}(0.9)
\end{aligned}
(\#eq:geneticexample)
\end{equation}

(ref:results-penalized-mean-caption) Results of segmentation by aplying the likelihood function defined in \@ref(eq:penalizedmeanlikelihood) to a sample of size 10 of the model defined in \@ref(eq:geneticexample).

(ref:eq-penalizedmeanlikelihood) \@ref(eq:penalizedmeanlikelihood)

(ref:eq-geneticexample) \@ref(eq:geneticexample)

```{r results-penalized-mean}
mean_likelihood <- function(X) {
  mean_value <- mean(X, na.rm = T)
  if (is.na(mean_value)) {
    0
  } else {
    -sum((X - mean_value)^2)
  }
}

penalized_mean_likelihood <- function(X) mean_likelihood(X) - 1

make_segment <- function(n, p) {
  matrix(rbinom(100 * n, 1, p), nrow = 100)
}

D_genetic <- cbind(
  make_segment(5, 0.9),
  make_segment(10, 0.1),
  make_segment(5, 0.9)
)

segment(
  D_genetic,
  likelihood = penalized_mean_likelihood,
  algorithm = "hieralg"
) %>%
  print_results_table(
    caption="Results of segmentation by aplying the
      likelihood function defined in
      (ref:eq-penalizedmeanlikelihood) to a sample of size
      10 of the model defined in (ref:eq-geneticexample).
    ")
```


