# Simulations

In order to exemplify the utility of this package, a handful of hypothetical
data will be presented, together with a proposal of an appropriate likelihood
function that is expected to segment the data in an expected manner.

## Correlated random variables

Let $C_1$, $C_2$, ..., $C_n$ a sequence of $n$ vectors of data points, each with $m$
observations represented by $C_k = \{c_{k1},\ c_{k2},\ ...,\ c_{km}\}$, with
$1 \le k \le n$. Assume that each $C_k$ is a function of an unknown set of independent
random variables $X$, such that $C_k = f_k(X)$. When used together with a multivariate
likelihood [@park2017fundamentals], for which an implementation is provided in
the `segmentr` package as the `multivariate()` function, it's possible to identify
the set of $N_b$ change points
$b_1, ..., b_{N_b}$, with $b_0 = 1$ and $b_{N_b + 1} = n + 1$, such that all $f_k$, for
$b_i \le k \le b_{i+1}-1$ is a function of the same set of independent random variables $X$,
or each $1 \le i \le N_b$.

Let $X_1, ..., X_6$ be independent random variables and $C_1, ..., C_{15}$ be
defined as in \@ref(eq:examplecolumns) . It's possible to see the first segment $C_1, ..., C_5$
depend on $X_1, X_2$, the second segment $C_6, ..., C_{10}$ depend on $X_3, X_4$ and the third
and last segment $C_{11}, ..., C_{15}$ depend on $X_4, X_5$. Therefore, the change points would
be $b_1 = 6$ and $b_2 = 11$.

\begin{equation}
\begin{aligned}
C_1 & = X_1 \\
C_2 & = X_1 - X_2 \\
C_3 & = X_2 \\
C_4 & = X_1 + X_2 \\
C_5 & = X_1 \\
C_6 & = X_3 \\
C_7 & = X_3 - X_4 \\
C_8 & = X_4 \\
C_9 & = X_3 + X_4 \\
C_{10} & = X_3 \\
C_{11} & = X_5 \\
C_{12} & = X_5 - X_6 \\
C_{13} & = X_6 \\
C_{14} & = X_5 + X_6 \\
C_{15} & = X_5
\end{aligned}
(\#eq:examplecolumns)
\end{equation}

To illustrate, consider the following code, which defines $X_1, ..., X_6$ and $C_1, ..., C_{15}$ as specified in \@ref(eq:examplecolumns), and also consider the matrix $D$ as the column binding of the vectors $C_1, ..., C_{15}$.

(ref:sample-columns-caption) Sample values of the correlated random variables defined in \@ref(eq:examplecolumns).

```{r sample-columns}
n <- 100
X1 <- sample(1:2, n, replace = TRUE)
X2 <- sample(1:2, n, replace = TRUE)
X3 <- sample(1:2, n, replace = TRUE)
X4 <- sample(1:2, n, replace = TRUE)
X5 <- sample(1:2, n, replace = TRUE)
X6 <- sample(1:2, n, replace = TRUE)

C1 <- X1
C2 <- X1 - X2
C3 <- X2
C4 <- X1 + X2
C5 <- X1
C6 <- X3
C7 <- X3 - X4
C8 <- X4
C9 <- X3 + X4
C10 <- X3
C11 <- X5
C12 <- X5 - X6
C13 <- X6
C14 <- X5 + X6
C15 <- X5
D <- cbind(C1, C2, C3, C4, C5, C6, C7, C8,
           C9, C10, C11, C12, C13, C14, C15)
head(D) %>%
  kable(caption="Sample values of the correlated random
        variables defined in (ref:eq-examplecolumns).")
```

It's possible to identify the segments in the matrix $D$ with a penalized version of the
`multivariate` function included in the package.


(ref:eq-examplecolumns) \@ref(eq:examplecolumns)

```{r results-columns}
print_results_table <- function(results, caption) {
  index <- seq(length(results$segments))
  starts <- sapply(results$segments, head, 1)
  ends <- sapply(results$segments, tail, 1)
  header <- c(2)
  names(header) <- paste0(
    "Segments (total of",
    length(results$segments),
    ")"
  )
  tibble(
    `Segment no.`=index,
    `First index`=starts,
    `Last index`=ends
  ) %>%
    kable(caption=caption)
}

penalized = function(X) multivariate(X) - 2 ^ ncol(X)
segment(D, likelihood = penalized, algorithm = "exact") %>%
  print_results_table(
    caption="Segment tables as defined in
    (ref:eq-examplecolumns).")
```

Notice it's important to use a penalized likelihood version of the `multivariate` function because
the original function favors bigger segments. The answer provided by the `segment` package
in that scenario would actually be all the columns in the $D$ matrix.

```{r}
segment(D, likelihood = multivariate, algorithm = "exact") %>%
  print_results_table(
    caption="Results of segmentation usign a non-penalized
      multivariate likelihood")
```

## Segments with similar averages

Segmentr can be use to segment random variables $C_1, ..., C_n$ that represent
genetic data, encoded as zero for homozygosity and
one for heterozygosity. In this scenario a researcher might be interested in finding
segments of homozygosity in the data, so it's desireable to segment the data
according to the homogeneity of values within a segment of DNA.

It's possible to use `segmentr` to find segments
that maximize the homogeneity of a segment with a "mean" likelihood function, i.e. a
function that maximizes the likelihood of segments whose elements approximate the
segment average. One such function is defined in \@ref(eq:mean-likelihood).

Therefore, it's possible to have `segmentr` find the change points in the data using
the adequate functions. It's necessary to penalize the likelihood function because
single columns segments actually maximize the function, which is not desirable.
Therefore, a constant penalty is included in order to penalize too many segments
as the final output.

Consider the simple example defined in \@ref(eq:geneticexample), in which
$X_i$ for $i \in \{1, \dots, 20\}$ represent each a column indexed by $i$ of a
data set $D$, and $\text{Bern(p)}$ representing the Berlinlli distribution wit
probability $p$. A penalized version of \@ref(eq:mean-likelihood) is defined in
\@ref(eq:penalizedmeanlikelihood), with the goal of punishing single column
segments, which would be the mathematical maximum of segmentation. By applying
the likelihood function defined in \@ref(eq:penalizedmeanlikelihood) to
the data set defined in \@ref(eq:geneticexample), the results obtained are
shown in Table \@ref(tab:results-penalized-mean).

\begin{equation}
L_{\mu}(X)=-\sum_i(x_i-E[X])^2 \;\bigg|\; x_i \in X
(\#eq:mean-likelihood)
\end{equation}

\begin{equation}
P_{L_{\mu}}(X)=L_{\mu}(X)-1
(\#eq:penalizedmeanlikelihood)
\end{equation}

\begin{equation}
\begin{aligned}
X_{1} & \sim \text{Bern}(0.9) \\
X_{2} & \sim \text{Bern}(0.9) \\
X_{3} & \sim \text{Bern}(0.9) \\
X_{4} & \sim \text{Bern}(0.9) \\
X_{5} & \sim \text{Bern}(0.9) \\
X_{6} & \sim \text{Bern}(0.1) \\
X_{7} & \sim \text{Bern}(0.1) \\
X_{8} & \sim \text{Bern}(0.1) \\
X_{9} & \sim \text{Bern}(0.1) \\
X_{10} & \sim \text{Bern}(0.1) \\
X_{11} & \sim \text{Bern}(0.1) \\
X_{12} & \sim \text{Bern}(0.1) \\
X_{13} & \sim \text{Bern}(0.1) \\
X_{14} & \sim \text{Bern}(0.1) \\
X_{15} & \sim \text{Bern}(0.1) \\
X_{16} & \sim \text{Bern}(0.9) \\
X_{17} & \sim \text{Bern}(0.9) \\
X_{18} & \sim \text{Bern}(0.9) \\
X_{19} & \sim \text{Bern}(0.9) \\
X_{20} & \sim \text{Bern}(0.9)
\end{aligned}
(\#eq:geneticexample)
\end{equation}

(ref:results-penalized-mean-caption) Results of segmentation by aplying the likelihood function defined in \@ref(eq:penalizedmeanlikelihood) to a sample of size 10 of the model defined in \@ref(eq:geneticexample).

(ref:eq-penalizedmeanlikelihood) \@ref(eq:penalizedmeanlikelihood)

(ref:eq-geneticexample) \@ref(eq:geneticexample)

```{r results-penalized-mean}
mean_likelihood <- function(X) {
  mean_value <- mean(X, na.rm = T)
  if (is.na(mean_value)) {
    0
  } else {
    -sum((X - mean_value)^2)
  }
}

make_segment <- function(n, p) {
  matrix(rbinom(100 * n, 1, p), nrow = 100)
}

D <- cbind(
  make_segment(5, 0.9),
  make_segment(10, 0.1),
  make_segment(5, 0.9)
)

segment(
  D,
  likelihood = function(X) mean_likelihood(X) - 1,
  algorithm = "hieralg"
) %>%
  print_results_table(
    caption="Results of segmentation by aplying the
      likelihood function defined in
      (ref:eq-penalizedmeanlikelihood) to a sample of size
      10 of the model defined in (ref:eq-geneticexample).
    ")
```


