# Considerations on performance

```{r}
source("helper.R")
```

This code shows a handful of attempts that were made in actually trying to execution
time of the algorithms implemented by the package.

## Benchmark the code

The package `microbenchmark` allows us to execute the code and compare different run-times.
Therefore it's going to be used in this package to compare the performance of different
code snippets.

```{r, message=FALSE}
library(microbenchmark)
```

## Native Code

In the algorithms chapter, we discussed the time complexity of the different types of
algorithms. In that scenario, the most relevant variable is the number of columns
in the data set, which dictates whether the computation of the exact algorithm will
be prohibitive or not. However, the number of samples does influence the computation
time linearly. Given the most performed operation in any of the algorithms described
in the chapter is the likelihood function, it's in essence the bottleneck of the
whole computation. Therefore, real performance gains can be obtained depending on
the performance of the application.

In R statistical programming, it's common place to use the functions provided by the
programming environment, which are usually performant algorithms implemented in compiled
programming languages such as C. However, there are times the programmer must implement
a custom function in R, which in turn is a dynamically, and typically slower execution
environment.

Therefore, in a first attempt to make the code execution run faster, and compare different
results, Segmentr implements the [multivariate()] likelihood function in compiled C++ programming language,
as well as it implements [r_multivariate()] which implements the same algorithm in the R
programming environment.

```{r benchmark-native}
data <- makeRandom(20, 100)

bench <- microbenchmark(
  segment(data,
          likelihood = multivariate,
          algorithm = "exact"),
  segment(data,
          likelihood = r_multivariate,
          algorithm = "exact"),
  segment(data,
          likelihood = multivariate,
          algorithm = "hierarchical"),
  segment(data,
          likelihood = r_multivariate,
          algorithm = "hierarchical"),
  times = 1
)

print_benchmark(
  bench,
  caption="Execution time comparison
    between native C++ and interpreted R code"
)
```

In the scenario above, we notice performance improvements in the native [multivariate()]
function, with the biggest performance improvement taking place in the `exact` algorithm
results. Therefore, whenever the resources are available, it's recommended to implement
the likelihood function directly in a native programming language such as C++. We recommend
the RCpp package, which makes it easy to implement functions which interface easily with
the R programming environment.

## Parallelization

Given the increasing trend of parallel programming in the past few years, it's natural
to try to take advantage of that by parallelizing the algorithm execution. In the
algorithms implemented in this paper, parallelization is a bit complicated because
a lot of the execution steps in the algorithm iterations depend on values computed
in previous steps, and the golden rule of parallel computing is that faster results
are generally obtained when computation steps are independent from one another, i.e.
no dependency between the task, hence they can be performed in parallel. Yet, there are
a few steps of the computation that can be performed in parallel, and its parallelization
was implemented with the help of the `foreach` package.

Therefore, whether or not the computation will be performed depends on whether there is
a parallel cluster registered in the R session with packages such as `doMC`, and whether
the `allow_parallel` argument is true or false. With that in mind, a few comparisons are evaluated.

```{r benchmark-parallel-narrow}
data <- makeRandom(100, 10)
doMC::registerDoMC(4)
bench <- microbenchmark(
  segment(data,
          likelihood = multivariate,
          algorithm = "exact",
          allow_parallel = FALSE),
  segment(data,
          likelihood = multivariate,
          algorithm = "exact",
          allow_parallel = TRUE),
  segment(data,
          likelihood = multivariate,
          algorithm = "hierarchical",
          allow_parallel = FALSE),
  segment(data,
          likelihood = multivariate,
          algorithm = "hierarchical",
          allow_parallel = TRUE),
  times = 1
)
print_benchmark(
  bench,
  caption="Execution time comparison
    between parralel and single-threaded computation
    with a data set of 100 samples and 10 columns"
)
```

Analyzing the results above, we notice good improvements, especially for the "exact" algorithm.
However, in a situation in which the number of rows isn't very large in comparison with the
number of columns, the results of the parallel computation are a lot more modest.
Consider the following case.

```{r benchmark-parallel-wide}
data <- makeRandom(5, 100)
doMC::registerDoMC(4)
bench <- microbenchmark(
  segment(data,
          likelihood = multivariate,
          algorithm = "exact",
          allow_parallel = FALSE),
  segment(data,
          likelihood = multivariate,
          algorithm = "exact",
          allow_parallel = TRUE),
  segment(data,
          likelihood = multivariate,
          algorithm = "hierarchical",
          allow_parallel = FALSE),
  segment(data,
          likelihood = multivariate,
          algorithm = "hierarchical",
          allow_parallel = TRUE),
  times = 1
)

print_benchmark(
  bench,
  caption="Execution time comparison
    between parralel and single-threaded computation
    with a data set of 5 samples and 100 columns"
)
```

In the result above, parallelization achieved the opposite effect, actually making the
computation slower, as there is now a lot of work done synchronizing necessary information
between all the processes in the cluster. That takes more time than is saved by running
the algorithm in parallel.

Taking all of that in consideration, the decision of whether or not to use parallelization
relies on the user after evaluating the nature of the data set. For data sets with few samples
and many columns, parallelization is not recommended. In the case there are lots of samples,
parallelization might give better computation times.

An alternative not developed in this project would be to use a genetic algorithm approach to
the minimal set estimation, which is a probabilistic process that converges to the result
after enough computation time have passed. The main advantage of such method is that computations
can be more easily parallelized. Such development might be explored in future developments of
this package.
