# Accuracy of algorithms

```{r}
source("helper.R")
```

Given different solutions obtained by the `segment` function, it's often necessary to compare them with each other. Therefore, it's useful to define a distance measure to compare two sets of estimated change points.

## The Hausdorff distance

As defined in [@topology], the Haurdorff distance is a measure of
how far two subsets from a metric space are from one another. It's
defined as the biggest of all the distances from a point in one
set to the closest point in the other set, which is expressed
mathematically by the equation defined in
\@ref(eq:hausdorffformula), in which $X$ and $Y$ represent sets
given as inputs to the function, $d(x, y)$ represents a distance
function defined in the metric space $S$ such that $x, y \in S$.

\begin{equation}
d_H(X,Y) = \max\left\{\,\sup_{x \in X} \inf_{y \in Y} d(x,y)\, ,\, \sup_{y \in Y} \inf_{x \in X} d(x,y)\,\right\}
(\#eq:hausdorffformula)
\end{equation}

Therefore, considering that a set of change points is a subset of
the indices of the column indices in a data set, if we define the
distance $d$ as the absolute value of two given numbers, we can
define the distance between two change points as the Hausdorff
distance between change point sets. One interesting application
of this is that it can be used to measure the distance of
an estimated set of change points to a reference solution.

## Accuracy of estimates for the Berlin dataset

In Chapter \@ref(real-data-examples) the data set of weather temperatures in
Berlin was presented, and there was a walk through on different ways to use
Segmentr to find a good estimate set of segments for it. Table
\@ref(tab:berlinexpectedtable) lists the considered the ideal solution to the
problem, at the same time as different solutions with different tables were
found in Tables \@ref(tab:resultsnonpenalized), \@ref(tab:resultsautopenalized)
and \@ref(tab:downresautopenalized). Considering this, the
Hausdorff distance can be used to measure how far each estimate is from
the ideal solution described in Table \@ref(tab:berlinexpectedtable).
In Table \@ref(tab:hausdorff-real-data-comparison) we can see the results
of the comparison, in which we can see the distance of each estimate
calculated.


```{r}
hausdorff <- function(set_a, set_b) {
  distance <- Vectorize(function(p1, p2) abs(p1 - p2))

  distance_1_to_2 <- function(set_1, set_2) {
    max(sapply(set_1, function(p1) min(distance(p1, set_2))))
  }

  max(distance_1_to_2(set_a, set_b), distance_1_to_2(set_b, set_a))
}

segment_distance <- function(changepoints1, changepoints2) {
  hausdorff(c(1, changepoints1), c(1, changepoints2))
}
```

(ref:table-berlin-expected) \@ref(tab:berlinexpectedtable)

(ref:table-non-penalized) \@ref(tab:resultsnonpenalized)

(ref:table-auto-penalized) \@ref(tab:resultsautopenalized)

(ref:table-adjusted-penalized) \@ref(tab:resultsadjustedpenalized)

(ref:table-rescaled) \@ref(tab:resultsrescaled)

```{r hausdorff-real-data-comparison}
deviation <- partial(
  segment_distance,
  changepoints2=expected_berlin$changepoints
)

tribble(
  ~`Results referenced`,
    ~`Estimated Changepoints`,
      ~`Hausdorff Distance to Ideal`,
  "Table (ref:table-berlin-expected)",
    comma_format(expected_berlin$changepoints),
      deviation(expected_berlin$changepoints),
  "Table (ref:table-non-penalized)",
    comma_format(results_non_penalized$changepoints),
      deviation(results_non_penalized$changepoints),
  "Table (ref:table-auto-penalized)",
    comma_format(results_auto_penalized$changepoints),
      deviation(results_auto_penalized$changepoints),
  "Table (ref:table-adjusted-penalized)",
    comma_format(results_adjusted_penalized$changepoints),
      deviation(results_adjusted_penalized$changepoints),
  "Table (ref:table-rescaled)",
    comma_format(results_rescaled$changepoints),
      deviation(results_rescaled$changepoints)
) %>% kable(
  caption="Hausdorff distance for different
  segmentation attepmts over the Berlin weather data set"
) %>% column_spec(2, width="15em")
```

In Table \@ref(tab:hausdorff-real-data-comparison) we can see the distance to
the ideal is gradually decreased more adequate parameters are used to compute
the segment estimates for the segmentation estimate. This improvement reflects
the progress that can be observed in gradually in Figures
\@ref(fig:berlin-non-penalized), \@ref(fig:berlin-auto-penalized),
\@ref(fig:berlin-adjusted-penalized) and \@ref(fig:plot-results-rescaled).

## Accuracy of algorithms using different algorithms

We want to measure how well the different algorithms provided in the Segmentr
package find the segments in the example correlated columns simulated data set
described in \@ref(eq:examplecolumns). Since the simulation was arbitrarily
built, we know what the exact solution to the segmentation problem is.


```{r hausdorff-multivariate-comparison}
expected_multivariate_example <- c(6, 11)

deviation <- partial(
  segment_distance,
  changepoints2=expected_multivariate_example
)

exact_multivariate_changepoints <- segment(
  D_example,
  likelihood = penalized_multivariate,
  algorithm = "exact"
)$changepoints

hieralg_multivariate_changepoints <- segment(
  D_example,
  likelihood = penalized_multivariate,
  algorithm = "hierarchical"
)$changepoints

hybrid_multivariate_changepoints <- segment(
  D_example,
  likelihood = penalized_multivariate,
  algorithm = "hybrid"
)$changepoints

hybrid_multivariate_changepoints_threshold <- segment(
  D_example,
  likelihood = penalized_multivariate,
  algorithm = "hybrid",
  threshold = 4
)$changepoints

tribble(
  ~`Description`,
    ~`Changepoints`,
      ~`Hausdorff Distance`,
  "Expected solution",
    comma_format(expected_multivariate_example),
      deviation(expected_multivariate_example),
  "Exact algorithm estimate",
    comma_format(exact_multivariate_changepoints),
      deviation(exact_multivariate_changepoints),
  "Hierarchical algorithm estimate",
    comma_format(hieralg_multivariate_changepoints),
      deviation(hieralg_multivariate_changepoints),
  "Hybrid algorithm estimate with threshold 50",
    comma_format(hybrid_multivariate_changepoints),
      deviation(hybrid_multivariate_changepoints),
  "Hybrid algorithm estimate with threshold 4",
    comma_format(hybrid_multivariate_changepoints_threshold),
      deviation(hybrid_multivariate_changepoints_threshold)
) %>%
  kable(caption="Comparison of solutions of different algorithms
to segmenting the data set described in (ref:eq-examplecolumns),
measuring how far each solution is from the ideal
solution using the Hausdorff distance.")
```

Therefore, in Table \@ref(tab:hausdorff-multivariate-comparison) we can see the
comparison of different algorithms with different parameters. We notice the
"exact" algorithm manages to properly match the expected change points
solution, whereas the "hierarchical" algorithm find an extra segment, which
causes the Hausdorff distance to be bigger than zero. The two "hybrid"
algorithm cases are interesting in the sense it that it shows how to algorithm
works by either adopting the exact or the hierarchical algorithm depending on
the threshold and the size of the segment being analyzed. When the threshold
argument is large, it applies the exact algorithm to the data set, whereas when
the threshold is small, it applies the hierarchical algorithm instead.

Consider now the simulated data set of segments with similar averages described
in equation (ref:eq-geneticexample). We do the same algorithm comparison with
this data set in Table \@ref(tab:hausdorff-mean-comparison). In contrast with
the previous experiment, all the algorithms manage to find the expected solution
to the problem.

```{r hausdorff-mean-comparison}
expected_mean_example <- c(6, 16)

deviation <- partial(
  segment_distance,
  changepoints2=expected_mean_example
)

exact_mean_changepoints <- segment(
  D_genetic,
  likelihood = penalized_mean_likelihood,
  algorithm = "exact"
)$changepoints

hieralg_mean_changepoints <- segment(
  D_genetic,
  likelihood = penalized_mean_likelihood,
  algorithm = "hierarchical"
)$changepoints

hybrid_mean_changepoints <- segment(
  D_genetic,
  likelihood = penalized_mean_likelihood,
  algorithm = "hybrid"
)$changepoints

hybrid_mean_changepoints_threshold <- segment(
  D_genetic,
  likelihood = penalized_mean_likelihood,
  algorithm = "hybrid",
  threshold = 4
)$changepoints

tribble(
  ~`Description`,
    ~`Changepoints`,
      ~`Hausdorff Distance`,
  "Expected solution",
    comma_format(expected_mean_example),
      deviation(expected_mean_example),
  "Exact algorithm estimate",
    comma_format(exact_mean_changepoints),
      deviation(exact_mean_changepoints),
  "Hierarchical algorithm estimate",
    comma_format(hieralg_mean_changepoints),
      deviation(hieralg_mean_changepoints),
  "Hybrid algorithm estimate with threshold 50",
    comma_format(hybrid_mean_changepoints),
      deviation(hybrid_mean_changepoints),
  "Hybrid algorithm estimate with threshold 4",
    comma_format(hybrid_mean_changepoints_threshold),
      deviation(hybrid_mean_changepoints_threshold)
) %>%
  kable(caption="Comparison of solutions of different algorithms
to segmenting the data set described in (ref:eq-geneticexample),
measuring how far each solution is from the ideal
solution using the Hausdorff distance.")
```

