# (APPENDIX) Appendix {-}

# Package Usage

The purpose of this work is too have the `segmentr` to be as accessible as possible.
Because of that, extra care was put into making it easy to install and use.

## Instalation

Given the algorithms and their applications discussed in [@base-paper],
the Segmentr package for R is proposed to help researchers segment their
data sets. Installation can be done using the default `install.packages` command,
like shown below.

```{r, echo=TRUE, eval=FALSE}
install.packages("segmentr")
```

## Usage

The package can be used with [segment()]. It takes a `data` argument containing a bi-dimensional matrix
in which the rows represent different samples and the columns represent the comprehension of the data set
we wish to segment. The function also accepts a `algorithm` argument, which can be `exact`, `hierarchical`
or `hybrid`, that specifies the type of algorithm will be used when exploring the data set. Finally, it's
also necessary to specify a `likelihood` function as argument, as it is used to compare and pick segments
that are better fit according to the given likelihood's chosen criteria.

```{r, echo=TRUE}
require("segmentr")

data <- rbind(
  c(1, 1, 0, 0, 0, 1023, 134521, 12324),
  c(1, 1, 0, 0, 0, -20941, 1423, 14334),
  c(1, 1, 0, 0, 0, 2398439, 1254, 146324),
  c(1, 1, 0, 0, 0, 24134, 1, 15323),
  c(1, 1, 0, 0, 0, -231, 1256, 13445),
  c(1, 1, 0, 0, 0, 10000, 1121, 331)
)

segment(
  data,
  algorithm = "exact",
  likelihood = function(X) multivariate(X) - 0.01*exp(ncol(X))
)
```

## Tutorial

In order to learn how to use the package to
analyze data, a vignette version of Chapter
\@ref(real-data-examples) is provided, such that
it can be used to follow the examples and learn
how to use the package.

```{r eval=FALSE}
vignette("segmentr")
```
