---
title: "An R Package for Change Point Detection"
title-formatted: "An R Package for \\linebreak Change Point Detection"
main-author: "Thales Mello"
advisor: "Florencia Leonardi"
author:
  - "Thales Mello"
  - "Florencia Leonardi"
valuation-committee:
  - "Profa. Dra. Florencia Leonardi"
  - "Prof. Dr. André Fujita"
  - "Prof. Dr. Jesús Enrique Garcia"
masters-program: "Mestrado em Estatística"
date: "São Paulo, 16 de Abril de 2019"
site: bookdown::bookdown_site
final-submission: true
presentation-date: "23 de Abril de 2019"
output: bookdown::gitbook
documentclass: book
classoption: twoside
bibliography: ["citations.bib"]
biblio-style: plainnat-ime.bst
link-citations: yes
paperize: a4
fontsize: 11pt
abstract-text: |
  \chapter*{Abstract}

  \noindent Mello, Thales. \textbf{An R Package for Change Point Detection}.
  2019. 61 f.
  Dissertação (Mestrado) - Instituto de Matemática e Estatística,
  Universidade de São Paulo, São Paulo, 2010.

  This dissertation explores the development of a package for the R programming
  language that allows the user to detect change points in a given data set.
  The change points are estimated using a segment cost function, which is
  chosen to suit the type of problem analyzed. The use cases mentioned in this
  work are (i) segments as blocks of independent discrete variables; and (ii)
  segments of random variables with a common regression parameter, for which a
  case of linear regression blocks and a case of Bernoulli distribution blocks
  are exemplified in this work.

  This work also describes the search algorithms implemented by the
  package (exact, hierarchical and mixed), each one having characteristics of
  performance and quality of the results distinct from the others.

  Finally, we explore some use-cases for the package, as well as some
  attempts at performance optimization, exploring the implementation of
  critical sections of code in low-level programming languages, as try to
  make use of parallel computing.

  \noindent \textbf{Keywords:} segment, change-point, cost, R-package.

resumo-text: |
  \chapter*{Resumo}

  \noindent Mello, Thales. \textbf{Um pacote R para detectar pontos de mudança}.
  2019. 61 f.
  Dissertação (Mestrado) - Instituto de Matemática e Estatística,
  Universidade de São Paulo, São Paulo, 2010.

  Esta dissertação explora o desenvolvimento de um pacote da linguagem de
  programação R com o objetivo de detectar pontos de mudança em um conjunto
  de dados a ser analisado. Os pontos de mudança são estimados com base em
  uma função de custo de segmentos, a qual é escolhida de acordo
  com o tipo de problema que se deseja analisar. Os casos de uso abordados
  neste trabalho são (i) segmentos como blocos independentes de variáveis
  aleatória discretas e (ii) segmentos como blocos de variáveis aleatórias
  com mesmo parâmetro de regressão, sendo demonstrado um exemplo com
  regressão linear e outro com a média dos valores numéricos do segmento.

  Neste trabalho são descritos os algorítmos de busca implementados pelo
  pacote (exato, hierárquico e misto), em que cada um possui características
  de performance e exatidão das estimativas próprios de cada algoritmo.

  Por fim, são exploradas algumas aplicações de uso do pacote, bem como
  algumas tentativas de otimização de performance, explorando
  implementação de trechos críticos do código em linguagens de baixo
  nível, bem como se valendo também do uso de computação paralela.

  \noindent \textbf{Palavras-chave:} segment, change-point, cost, R-package.
acknowledgements-text: |
  \chapter*{Acknowledgements}

  I thank Bruno M. Castro and Florencia Leonardi, for their work in the paper
  that inspired this project. I also thank specially Handley Wickham, for his
  work in several R packages and his tutorial on the development of R packages,
  and Yihui Xie for his work in the bookdown package. Without their
  contributions to the R community, none of this would be possible.

---

```{r message=FALSE, echo=FALSE}
library(knitr)
library(kableExtra)
library(segmentr)
library(tibble)
library(magrittr)
library(purrr)
knitr::opts_chunk$set(
  cache=TRUE,
  echo=FALSE
)
```

```{r, code = readLines("helper.R")}
```

<!-- Macros used in this project -->
\newcommand{\X}{\bf X}
\newcommand{\x}{\mathbf x}
\newcommand{\Xs}{\mathbf X}
\renewcommand{\P}{\mathbb P}
\newcommand{\E}{\mathbb E}
\newcommand{\argmax}{\arg\max}
\newcommand{\pen}{\text{pen}}
\newcommand{\p}{\text{p}}
\newcommand{\wQ}{\widetilde Q}
\newcommand{\wP}{\widehat\Pr}
\newcommand{\M}{\mathcal M}
\newcommand{\rs}{r\!:\!s}
\newcommand{\1}{\mathbf 1}
\newcommand{\R}{\mathbb R}
\newcommand{\Z}{\mathbb Z}
\newcommand{\lik}{\mathcal{L}}
\newcommand{\lt}{<}
\newcommand{\gt}{>}

