---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# segmentr

[![Travis-CI Build Status](https://travis-ci.org/thalesmello/segmentr.svg?branch=master)](https://travis-ci.org/thalesmello/segmentr)

Given a likelihood provided by the user, this package applies it to a given
matrix dataset in order to find change points in the data that maximize the sum
of the likelihoods of all the segments.

This package provides a handful of algorithms with different time complexities
and assumption compromises so the user is able to choose the best one for the
problem at hand.

## Installation

You can install segmentr from GitHub using:

```{r gh-installation, eval = FALSE}
install.packages("segmentr")
```

## Example

Sample code using the package to identify change points in the segments' averages:

```{r example}
require(segmentr)

make_segment <- function(n, p) matrix(rbinom(100 * n, 1, p), nrow = 100)
data <- cbind(make_segment(5, 0.1), make_segment(10, 0.9), make_segment(2, 0.1))
mean_lik <- function(X) abs(mean(X) - 0.5) * ncol(X)^2
segment(data, likelihood = mean_lik, algorithm = "hieralg")
```

For an in depth step-by-step, please check `vignette("segmentr")`.

## Acknowledgments

This package is part of a Master's degree research thesis at [IME-USP](https://www.ime.usp.br/),
with [Florencia Leonardi][florencia] as thesis adviser.

The algorithms in this package are based on a [paper][basis_paper] by Bruno M. de Castro and [Florencia Leonardi][florencia].

The `berlin` sample dataset was provided by [© Deutscher Wetterdienst][dwd] and put together with the [`rdwd`][rdwd] package by [Berry Boessenkool][brry]. Check [`make_berlin.R`][make_berlin] for the script that builds the dataset.

[basis_paper]: https://arxiv.org/abs/1501.01756
[florencia]: https://www.ime.usp.br/~leonardi/Home_page_of_Florencia_G._Leonardi/Home_Page.html
[dwd]: https://dwd.de/
[rdwd]: https://CRAN.R-project.org/package=rdwd
[brry]: https://github.com/brry
[make_berlin]: https://github.com/thalesmello/segmentr/blob/master/make_berlin.R
